---
title: "Introduction to dblinkR"
author: "Neil Marchant and Rebecca C. Steorts"
date: "18 March 2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to dblinkR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

We present an example from Marchant, Kaplan, Elazar, Rubinstein, and Steorts (2020), "d-blink: Distributed End-to-End Bayesian Entity Resolution." We will be using a dataset that was formally in the deprecated RecordLinkage package in R, known as the RLdata500 data set. This data set contains first and last name and full date of birth. 

The dblinkR package performs joint blocking and entity resolution from multiple databases using the methods outlined in the paper above. 

Our goals include:

- Loading Required Packages and Connecting to Spark 
- Presenting the RLdata500 dataset with summary information.
- Illustrating how we can format the RLdata500 dataset to work with the dblinkR package
- Running Inference for dblinkR
- Looking a Diagnostics Results 
- Performing Evaluation Metrics 

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=7, 
  fig.height=5
)
```


## Loading Required Packages

We begin by loading the required packages. Note that the `dblinkR` package 
must be loaded *after* `sparklyr`.
```{r packages, message=FALSE}
# For running dblink
library(sparklyr)
library(dblinkR)

# For generating trace plots
library(stringr)
library(dplyr)
library(tidyr)
library(ggplot2)

# For generating credible interval plots
library(tidybayes)
```

## Connecting to Spark
In this vignette, we will run `dblink` on a local instance of Spark with 2 
cores. This is sufficient for testing purposes, however for larger data sets, 
we recommend connecting to a Spark deployment. The `sparklyr` 
[documentation](https://spark.rstudio.com/deployment/) contains information 
about connecting to Spark deployments.

```{r spark-connect, message=FALSE}
sc <- spark_connect(master = "local[2]", version = "2.4.3")
spark_context(sc) %>% invoke("setLogLevel", "WARN")
```

`dblink` requires a location on disk (includes HDFS) to save diagnostics, 
posterior samples, and the state of the Markov chain. We refer to this location 
as the *project path*, and set it to the current working directory below. 
In addition, we must also specify a location on disk to save Spark checkpoints. 
```{r set-paths, message=FALSE}
projectPath <- paste0(getwd(), "/") # working directory
spark_set_checkpoint_dir(sc, paste0(projectPath, "checkpoints/"))
```

Now, that we have loaded packages that are needed and connected to spark, we can proceed with the rest of the tutorial! 

## Understanding the RLdata500 dataset

The RLdata500 data consists of 500 records with 10 percent duplication. Thus, there are 450 unique individuals. There is full information on each record containing first name, last name, and full date of birth. We also load in the unique identifiers, so that evaluations can be performed later on. 

We provide the first few lines of the data below.

```{r load-data}
records <- RLdata500
records['rec_id'] <- seq_len(nrow(records))
records <- lapply(records, as.character) %>% as.data.frame(stringsAsFactors = FALSE)
# inspect RLdata500
head(RLdata500)
```


## dblink parameters

Here, we specify the parameters for the `dblink` model. 

We treat the name-related attributes as string-type with a Levenshtein similarity 
function, and the date-related attributes as categorical (with a constant 
similarity function). We also place a Beta prior on the distortion probability 
for each attribute. 

```{r set-attribute-parameters}
distortionPrior <- BetaRV(1, 50)
levSim <- LevenshteinSimFn(threshold = 7, maxSimilarity =  10)

attributeSpecs <- list(
  fname_c1 = Attribute(levSim, distortionPrior),
  lname_c1 = Attribute(levSim, distortionPrior),
  by = CategoricalAttribute(distortionPrior),
  bm = CategoricalAttribute(distortionPrior),
  bd = CategoricalAttribute(distortionPrior)
)
```

The Beta prior on the distortion probabilities favours low distortion, as 
shown below.

```{r plot-distortion-prior}
tibble(prob = seq(from=0, to=1, by=0.01)) %>%
  mutate(density = dbeta(prob, distortionPrior@shape1, distortionPrior@shape2)) %>%
  ggplot(aes(x = prob, y = density)) + geom_line() + 
  labs(title = "Prior distribution on the distortion probabilities", 
       x = "Distortion probability", y = "Density")
```

We can also specify the size of the latent entity population. A larger 
population size favours under-linkage, while a smaller population 
size favours over-linkge. By default, the population size is set to the 
number of records in the data. This is specified by setting the population 
size to `NULL`:

```{r set-population-size}
populationSize <- NULL
```

Finally, we specify the type of blocking (or the partitioner). Since the data is small in this case, 
we can switch partitioning off by setting the partitioner to `NULL`. 
For larger data sets, we recommend using the `KDTreePartitioner`.

```{r set-partitioner}
partitioner <- NULL
```

## Running inference
In the previous section, we specified the model parameters. Now, we are ready to perform inference. 
The results will be saved in the project path.

```{r run-inference}
state <- initializeState(sc, records, attributeSpecs, recIdColname = 'rec_id',
                         partitioner = partitioner, populationSize = populationSize,
                         randomSeed = 1, maxClusterSize = 10L)

result <- runInference(state, projectPath, sampleSize = 1000,
                       burninInterval = 5000, thinningInterval = 10)
```

To load previously saved results, we can use the following function.

```{r, eval=FALSE}
#load("result.Rdata")
  #result <- loadResult(sc, projectPath)
```


## MCMC diagnostics
As one knows, it's important to review Markov chain Monte carlo (MCMC) diagnostics after performing 
inference. Such diagnostics saved in the `diagnostics.csv` file located in the project path. 
We can use the function below to load these into a local tibble.

```{r read-diagnostics}
diagnostics <- loadDiagnostics(sc, projectPath)
```

It's useful to examine trace plots, to assess the convergence.
```{r diagnostic-plots-1}
ggplot(diagnostics, aes(x=iteration, y=numObservedEntities)) + 
  geom_line() + 
  labs(title = 'Trace plot: # entity clusters', x = 'Iteration', 
       y = '# clusters')

diagnostics %>%
  select(iteration, starts_with("aggDist")) %>%
  gather(attribute, numDistortions, starts_with("aggDist")) %>%
  mutate(attribute = str_match(attribute, "^aggDist(.+)")[,2],
         numDistortions = numDistortions / nrow(records)) %>% 
  ggplot(aes(x=iteration, y=numDistortions)) +  
  geom_line(aes(colour = attribute)) + 
  labs(title = 'Trace plot: attribute distortion', x = 'Iteration', 
       y = '% distorted', colour = 'Attribute')

ggplot(diagnostics, aes(x=iteration, y=logLikelihood)) + 
  geom_line() + 
  labs(title = 'Trace plot: log-likelihood (unnormalized)', x = 'Iteration', 
       y = 'log-likelihood')
```

Additional diagnostics statistics can be computed from the samples of the 
linkage structure, which we refer to as the *linkage chain*. 

```{r diagnostic-plots-2}
linkageChain <- loadLinkageChain(sc, projectPath)

clustSizeDist <- clusterSizeDistribution(linkageChain)
ggplot(clustSizeDist, aes(x=iteration, y=frequency)) + 
  geom_line(aes(colour = clusterSize)) + 
  labs(title = 'Trace plot: cluster size distribution', x = 'Iteration', 
       y = 'Frequency', colour = 'Cluster size')

partSizes <- partitionSizes(linkageChain)
ggplot(partSizes, aes(x=iteration, y=size)) + 
  geom_line(aes(colour = partitionId)) + 
  labs(title = 'Trace plot: partition sizes', x = 'Iteration', y = 'Size', 
       colour = 'Partition')
```

## Evaluation

In this section, we consider many types of evaluation of our proposed methodology:

- the linkage quality
- the estimates posterior summary statistics (posterior mean, posterior variance, and credible intervals)
- the recall and precision

### Linkage quality
We compute a point estimate of the posterior linkage structure/clustering 
using the *shared most probable maximal matching sets method*. Note that we 
must use the `collect` function to retrieve the data from Spark into a local 
tibble.
```{r smpc}
predClusters <- dblinkR::sharedMostProbableClusters(linkageChain) %>% collect()
```

Next we assess the quality of the predicted linkage structure by 
computing pairwise precision and recall using functions from the 
[`exchangeableER`](https://github.com/ngmarchant/exchangeableER) package.
```{r pairwise-metrics, message=FALSE}
library(exchangeableER)
trueClusters <- exchangeableER::membershipToClusters(identity.RLdata500, ids = records$rec_id)
predMatches <- exchangeableER::clustersToPairs.Clusters(predClusters)
trueMatches <- exchangeableER::clustersToPairs.Clusters(trueClusters)
numRecords <- nrow(records)
conMat <- exchangeableER::confusionMatrix(predMatches, trueMatches, numRecords*(numRecords - 1)/2)
print(exchangeableER::pairwiseMetrics(conMat))
```

### Posterior estimates
We can use the saved posterior samples to infer unknown model parameters. 
First, we consider the number of unique entities present in the data. 
This summary statistic is stored in `diagnostics$numObservedEntities`.
Below we compute a point estimate based on the median, along with a 95\% 
highest density interval.

```{r}
hdiNumEntities <- median_hdih(diagnostics$numObservedEntities)
cat("The predicted number of unique entities in RLdata500 is ", 
    hdiNumEntities$x, ".\n", sep="")
cat("A 95% HDI is ", 
    sprintf("[%s,%s]", hdiNumEntities$xmin, hdiNumEntities$xmax), ".\n", sep="")
```

We can also visualize the estimated posterior distribution, while 
comparing to the ground truth (red), posterior median (blue) and 
95\% highest density interval (green). 

```{r posterior-plot}
ggplot(diagnostics) + 
  geom_histogram(aes(x=numObservedEntities), binwidth = 1) + 
  geom_vline(aes(xintercept = hdiNumEntities$x), color='blue') + 
  geom_vline(aes(xintercept = 450), color='red') + 
  geom_vline(aes(xintercept = hdiNumEntities$xmin), color = 'green') + 
  geom_vline(aes(xintercept = hdiNumEntities$xmax), color = 'green') + 
  labs(x = "# of entities", y = "Frequency", title = "Posterior number of observed entities")
```

Second, we consider the distortion level observed in each attribute, averaged 
across all of the records. This information is also stored in 
`diagnostics`. Below we plot a point estimate of the distortion level for each 
attribute based on the median, along with a 95\% highest density interval.

```{r posterior-attribute-distortion}
diagnostics %>%
  select(iteration, starts_with("aggDist")) %>%
  gather(attribute, numDistortions, starts_with("aggDist")) %>%
  transmute(attribute = str_match(attribute, "^aggDist(.+)")[,2],
            percDistorted = numDistortions / nrow(records)) %>% 
  group_by(attribute) %>% 
  ggplot(aes(y = attribute, x = percDistorted)) + 
    stat_pointintervalh(point_interval = median_hdi, .width = 0.95) + 
    labs(x = "Attribute", y = "Distortion level (%)", title = "Posterior distortion level by attribute")
```
