---
title: "Introduction to dblinkR"
author: "Neil Marchant and Rebecca C. Steorts"
date: "18 March 2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to dblinkR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this vignette, we demonstrate how to use dblinkR to perform Bayesian 
entity resolution. dblinkR implements a generative Bayesian model that 
jointly performs blocking and entity resolution of multiple databases as 
outlined in the following paper:

> Marchant, N. G., Steorts, R. C., Kaplan, A., Rubinstein, B. I. P., & Elazar, 
D. N. (2019). d-blink: Distributed End-to-End Bayesian Entity Resolution. 
arXiv preprint arXiv:1909.06039.

We will be using a dataset called `RLdata500` that was formerly available in 
the RecordLinkage R package (now deprecated). For convenience, we have 
included the `RLdata500` dataset in `dblinkR`.

An outline of the vignette is as follows:

1. We load the required packages and connect to Spark 
2. We introduce the RLdata500 dataset
3. We demonstrate how to set the d-blink model parameters
4. We perform inference using MCMC
5. We examine MCMC diagnostics
6. We assess the posterior linkage structure and other model parameters

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=7, 
  fig.height=5
)
```


## Loading Required Packages

We begin by loading the required packages. Note that the `dblinkR` package 
must be loaded *after* `sparklyr`.
```{r packages, message=FALSE}
# For running dblink
library(sparklyr)
library(dblinkR)

# For generating trace plots
library(stringr)
library(dplyr)
library(tidyr)
library(ggplot2)

# For generating credible interval plots
library(tidybayes)
```

## Connecting to Spark
In this vignette, we will run `dblink` on a local instance of Spark with 2 
cores. This is sufficient for testing purposes, however for larger data sets, 
we recommend connecting to a Spark deployment. The `sparklyr` 
[documentation](https://spark.rstudio.com/deployment/) contains information 
about connecting to Spark deployments.

```{r spark-connect, message=FALSE}
sc <- spark_connect(master = "local[2]", version = "2.4.3")
spark_context(sc) %>% invoke("setLogLevel", "WARN")
```

`dblink` requires a location on disk (includes HDFS) to save diagnostics, 
posterior samples, and the state of the Markov chain. We refer to this location 
as the *project path*, and set it to the current working directory below. 
In addition, we must also specify a location on disk to save Spark checkpoints. 
```{r set-paths, message=FALSE}
projectPath <- paste0(getwd(), "/") # working directory
spark_set_checkpoint_dir(sc, paste0(projectPath, "checkpoints/"))
```

Now, that we have loaded the required packages and connected to Spark, we can 
proceed with the rest of the tutorial! 

## Understanding the RLdata500 dataset

RLdata500 is a synthetic dataset that contains 500 personal records with 10 
percent duplication. Thus, there are 450 unique individuals represented in 
the data and 50 duplicate records. Each record contains the individual's 
full name and date of birth, however these attributes may be distorted. 
We add an additional column for record identifiers, which are used to specify 
the linkage structure later on.

```{r load-data}
records <- RLdata500
records['rec_id'] <- seq_len(nrow(records))
records <- lapply(records, as.character) %>% as.data.frame(stringsAsFactors = FALSE)
# inspect RLdata500
head(RLdata500)
```


## dblink parameters

Here, we specify the parameters for the `dblink` model. 

We treat the name-related attributes as string-type with a Levenshtein 
similarity function, and the date-related attributes as categorical (with a 
constant similarity function). We also place a Beta prior on the distortion 
probability for each attribute. 

```{r set-attribute-parameters}
distortionPrior <- BetaRV(1, 50)
levSim <- LevenshteinSimFn(threshold = 7, maxSimilarity =  10)

attributeSpecs <- list(
  fname_c1 = Attribute(levSim, distortionPrior),
  lname_c1 = Attribute(levSim, distortionPrior),
  by = CategoricalAttribute(distortionPrior),
  bm = CategoricalAttribute(distortionPrior),
  bd = CategoricalAttribute(distortionPrior)
)
```

The Beta prior on the distortion probabilities favours low distortion, as 
shown below.

```{r plot-distortion-prior}
tibble(prob = seq(from=0, to=1, by=0.01)) %>%
  mutate(density = dbeta(prob, distortionPrior@shape1, distortionPrior@shape2)) %>%
  ggplot(aes(x = prob, y = density)) + geom_line() + 
  labs(title = "Prior distribution on the distortion probabilities", 
       x = "Distortion probability", y = "Density")
```

We can also specify the size of the latent entity population. A larger 
population size favours under-linkage, while a smaller population 
size favours over-linkge. By default, the population size is set to the 
number of records in the data. This is specified by setting the population 
size to `NULL`:

```{r set-population-size}
populationSize <- NULL
```

Finally, we specify the partitioner which is used for blocking the entities 
to improve scalability. Since the data is small in this case, we can switch 
partitioning off by setting the partitioner to `NULL`. For larger data sets, 
we recommend using the `KDTreePartitioner` (please refer to the documentation).

```{r set-partitioner}
partitioner <- NULL
```

## Running inference
In the previous section, we specified the model parameters. Now, we are ready 
to perform inference. The results will be saved in the project path.

```{r run-inference}
state <- initializeState(sc, records, attributeSpecs, recIdColname = 'rec_id',
                         partitioner = partitioner, populationSize = populationSize,
                         randomSeed = 1, maxClusterSize = 10L)

result <- runInference(state, projectPath, sampleSize = 1000,
                       burninInterval = 5000, thinningInterval = 10)
```

To load results from a previous run, we can use the following function.

```{r, eval=FALSE}
#result <- loadResult(sc, projectPath)
```


## MCMC diagnostics

It's important to review convergence and mixing diagnostics when performing 
inference using Markov chain Monte Carlo (MCMC). Various diagnostics are saved 
in the `diagnostics.csv` file located in the project path. We can use the 
function below to load these into a local tibble.

```{r read-diagnostics}
diagnostics <- loadDiagnostics(sc, projectPath)
```

Below we provide trace plots of various summary statistics.

```{r diagnostic-plots-1}
ggplot(diagnostics, aes(x=iteration, y=numObservedEntities)) + 
  geom_line() + 
  labs(title = 'Trace plot: # entity clusters', x = 'Iteration', 
       y = '# clusters')

diagnostics %>%
  select(iteration, starts_with("aggDist")) %>%
  gather(attribute, numDistortions, starts_with("aggDist")) %>%
  mutate(attribute = str_match(attribute, "^aggDist(.+)")[,2],
         numDistortions = numDistortions / nrow(records)) %>% 
  ggplot(aes(x=iteration, y=numDistortions)) +  
  geom_line(aes(colour = attribute)) + 
  labs(title = 'Trace plot: attribute distortion', x = 'Iteration', 
       y = '% distorted', colour = 'Attribute')

ggplot(diagnostics, aes(x=iteration, y=logLikelihood)) + 
  geom_line() + 
  labs(title = 'Trace plot: log-likelihood (unnormalized)', x = 'Iteration', 
       y = 'log-likelihood')
```

Additional diagnostic statistics can be computed from the samples of the 
linkage structure, which we refer to as the *linkage chain*. 

```{r diagnostic-plots-2}
linkageChain <- loadLinkageChain(sc, projectPath)

clustSizeDist <- clusterSizeDistribution(linkageChain)
ggplot(clustSizeDist, aes(x=iteration, y=frequency)) + 
  geom_line(aes(colour = clusterSize)) + 
  labs(title = 'Trace plot: cluster size distribution', x = 'Iteration', 
       y = 'Frequency', colour = 'Cluster size')

partSizes <- partitionSizes(linkageChain)
ggplot(partSizes, aes(x=iteration, y=size)) + 
  geom_line(aes(colour = partitionId)) + 
  labs(title = 'Trace plot: partition sizes', x = 'Iteration', y = 'Size', 
       colour = 'Partition')
```

## Evaluation

In this section, we evaluate the predictions of our model and compare against 
the ground truth (provided with the synthetic data). 

We compute pairwise precision and recall using a point estimate of the 
linkage structure. We also inspect the posterior distribution over two 
important summary statistics:

* the number of unique individuals in the data, and
* the distortion level of each attribute.

### Linkage quality
We compute a point estimate of the posterior linkage structure/clustering 
using the *shared most probable maximal matching sets method* 
(Steorts et al., 2016). This method creates a globally consistent estimate 
that obeys transitivity constraints.

Note that we must use the `collect` function to retrieve the data from Spark 
into a local tibble.

```{r smpc}
predClusters <- sharedMostProbableClusters(linkageChain) %>% collect()
```

Next we assess the quality of the predicted linkage structure by 
computing pairwise precision and recall using functions from the 
[`clevr`](https://github.com/cleanzr/clevr) package.

```{r pairwise-metrics, message=FALSE}
library(clevr)
predMatches <- clusters_to_pairs(predClusters)
trueMatches <- membership_to_pairs(identity.RLdata500, elem_ids = records$rec_id)
numRecords <- nrow(records)
measures <- eval_report_pairs(trueMatches, predMatches, numRecords*(numRecords - 1)/2)
print(measures)
```

### Posterior estimates

We can use the saved posterior samples to infer unknown model parameters. 
First, we consider the number of unique entities present in the data. 
This summary statistic is stored in `diagnostics$numObservedEntities`.
Below we compute a point estimate based on the median, along with a 95\% 
highest density interval.

```{r}
hdiNumEntities <- median_hdi(diagnostics$numObservedEntities)
cat("The predicted number of unique entities in RLdata500 is ", 
    hdiNumEntities$y, ".\n", sep="")
cat("A 95% HDI is ", 
    sprintf("[%s,%s]", hdiNumEntities$ymin, hdiNumEntities$ymax), ".\n", sep="")
```

We can also visualize the estimated posterior distribution, while 
comparing to the ground truth (red), posterior median (blue) and 
95\% highest density interval (green). 

```{r posterior-plot}
ggplot(diagnostics) + 
  geom_histogram(aes(x=numObservedEntities), binwidth = 1) + 
  geom_vline(aes(xintercept = hdiNumEntities$y), color='blue') + 
  geom_vline(aes(xintercept = 450), color='red') + 
  geom_vline(aes(xintercept = hdiNumEntities$ymin), color = 'green') + 
  geom_vline(aes(xintercept = hdiNumEntities$ymax), color = 'green') + 
  labs(x = "# of entities", y = "Frequency", title = "Posterior number of observed entities")
```

Second, we consider the distortion level observed in each attribute, averaged 
across all of the records. This information is also stored in 
`diagnostics`. Below we plot a point estimate of the distortion level for each 
attribute based on the median, along with a 95\% highest density interval.

```{r posterior-attribute-distortion}
diagnostics %>%
  select(iteration, starts_with("aggDist")) %>%
  gather(attribute, numDistortions, starts_with("aggDist")) %>%
  transmute(attribute = str_match(attribute, "^aggDist(.+)")[,2],
            percDistorted = numDistortions / nrow(records) * 100) %>% 
  group_by(attribute) %>% 
  ggplot(aes(y = attribute, x = percDistorted)) + 
    stat_pointinterval(point_interval = median_hdi, .width = 0.95) + 
    labs(y = "Attribute", x = "Distortion level (%)", title = "Posterior distortion level by attribute")
```
